---
title: "Data Analysis"
author: "Claire Willing"
date: "2024-12-10"
output: html_document
---

### Read in libraries

```{r warning=FALSE}
# Function to install and load packages
install_and_load <- function(packages) {
  for (pkg in packages) {
    if (!requireNamespace(pkg, quietly = TRUE)) {
      if (pkg == "phyloseq") {
        if (!requireNamespace("BiocManager", quietly = TRUE)) {
          install.packages("BiocManager")
        }
        BiocManager::install(pkg)
      } else {
        install.packages(pkg)
      }
    }
    library(pkg, character.only = TRUE)
  }
}

# List of required packages
required_packages <- c(
  "phyloseq",  # For working with sequencing data
  "dplyr",     # For data manipulation
  "ggplot2",   # For plotting
  "readxl",    # For reading Excel files
  "writexl",   # For writing Excel files
  "tibble",    # For converting columns to row names
  "stringr",   # For string manipulation
  "viridis",   # For color palettes
  "vegan",     # For statistics
  "tidyr",     # For data org
  "knitr",
  "lme4",
  "ggeffects", 
  "cowplot"
)

# Install and load packages
install_and_load(required_packages)
```

### Load data files

Make sure to edit the name of your OTU table (eg STUDY_phyloseq_rare.rds)

```{r}
seqdata_RARE<-readRDS(
  "data/processed_data/phyloseq/STUDY_phyloseq_rare.rds")
seqdata_NOTRARE<-readRDS(
  "data/processed_data/phyloseq/STUDY_phyloseq_notrare.rds")
```

###Filter out plant reads
```{r}
seqdata_RARE <- subset_taxa(seqdata_RARE, Kingdom == "Fungi")
```

### Ordination (NMDS)

Using default of bray for distance (bray is default method for ordinate function)

```{r message=FALSE, warning=FALSE}

ORDtotalvals<-ordinate(seqdata_RARE, "NMDS")

ORD<-plot_ordination(seqdata_RARE, 
              ORDtotalvals, color="burn_temp_C", shape="burn_status")+
              theme_minimal()
ORD
```

### Quick barplot

```{r message=FALSE, warning=FALSE}
# Step 1: Merge samples by a grouping variable (e.g., "SampleType")
ps_merged <- seqdata_RARE %>%
  merge_samples("burn_temp_C")

# Step 2: Convert to relative abundance
ps_rel <- ps_merged %>%
  transform_sample_counts(function(x) x / sum(x))

# Step 3: Aggregate at the taxonomic level of interest (e.g., "Order")
ps_taxa <- ps_rel %>%
  tax_glom(taxrank = "Family")

# Step 4: Melt the phyloseq object for visualization
ps_melt <- ps_taxa %>%
  psmelt() %>%
  mutate(
    Sample = factor(Sample, levels = unique(Sample)),  # Ensure Sample is a factor
    Family = ifelse(Abundance < 0.02, "Other", Family)   # Group low-abundance taxa as "Other"
  )

##convert burn temp to factor
ps_melt$burn_temp_C <- factor(ps_melt$burn_temp_C, 
                                   levels = c(20, 45, 75, 105, 150), ordered = TRUE)

ps_melt$Sample <- factor(ps_melt$Sample, levels = unique(ps_melt$Sample[order(ps_melt$burn_temp_C)]))

##normalize abundance
ps_melt <- ps_melt %>%
  group_by(Sample) %>%
  mutate(RelativeAbundance = Abundance / sum(Abundance) * 100)

# Step 5: Create the bar plot
barplot <- ps_melt %>%
  ggplot(aes(x = Sample, burn_temp_C, y = RelativeAbundance, fill = Family)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(
    x = "Group",
    y = "Relative Abundance",
    fill = "Family") +
  theme_minimal() +
  scale_fill_viridis_d(option = "plasma")
    # Use viridis color scale

# Display the plot
print(barplot)
```

### Saving figures

Here is an example for saving figures for publication:

*If you don't yet have a directory for figures, you can make one here:*

```{bash eval=FALSE, engine.opts='-l', include=FALSE}
mkdir -p figures/exploratory
mkdir -p figures/publication
```

Now save this in your figures directory

```{r}
##Ordination
#ggsave("figures/exploratory/STUDY_bray_ordination.pdf", plot = ORD, width = 6, height = 4, dpi = 300)

##Barplot
#ggsave("figures/exploratory/STUDY_barplot_burn-status.pdf", plot = barplot, width = 6, height = 4, dpi = 300)
```

You can also use cowplot or other packages if you need to combine figures and add in figure letters

```{r}
combined_plot <- plot_grid(
  ORD, barplot,
  labels = c("A", "B"),  # Add figure letters
  label_size = 12,        # Adjust label size
  ncol = 2                # Arrange plots in 2 columns
)

combined_plot
```

Now save this combined figure in your figures directory

```{r}
#ggsave("figures/publication/Figure1_AB.pdf", plot = combined_plot, width = 6, height = 4, dpi = 300)
```
