---
title: "exploratory figures"
author: "Mira Ranganath"
date: "2025-03-08"
output: html_document
---

```{r setup, include=FALSE}
detach_package <- function(pkg, character.only = FALSE) {
  if (!character.only) pkg <- deparse(substitute(pkg))
  search_item <- paste0("package:", pkg)
  if (search_item %in% search()) {
    detach(search_item, unload = TRUE, character.only = TRUE)
  }
}
```

# Read in libraries
```{r warning=FALSE}
# Function to install and load packages
install_and_load <- function(packages) {
  for (pkg in packages) {
    if (!requireNamespace(pkg, quietly = TRUE)) {
      if (pkg == "phyloseq") {
        if (!requireNamespace("BiocManager", quietly = TRUE)) {
          install.packages("BiocManager")
        }
        BiocManager::install(pkg)
      } else {
        install.packages(pkg)
      }
    }
    library(pkg, character.only = TRUE)
  }
}

# List of required packages
required_packages <- c(
  "phyloseq",  # For working with sequencing data
  "dplyr",     # For data manipulation
  "ggplot2",   # For plotting
  "readxl",    # For reading Excel files
  "writexl",   # For writing Excel files
  "tibble",    # For converting columns to row names
  "stringr",   # For string manipulation
  "viridis",   # For color palettes
  "vegan",     # For statistics
  "tidyr",     # For data org
  "lme4",
  "ggeffects", 
  "cowplot",
  "here"
)

# Install and load packages
install_and_load(required_packages)
```

# Load data files
```{r}
seqdata_RARE<-readRDS(
  "data/processed_data/phyloseq/STUDY_phyloseq_rare.rds")
seqdata_NOTRARE<-readRDS(
  "data/processed_data/phyloseq/STUDY_phyloseq_notrare.rds")
```

## Filter out plant reads
```{r}
seqdata_RARE <- subset_taxa(seqdata_RARE, Kingdom == "Fungi")
```

##Creating new df with OTUS
```{r}
tax_mat <- tax_table(seqdata_RARE)

tax_mat <- tax_mat %>% 
  as.data.frame() %>% 
  select(1:7)

# Convert to character matrix to avoid factor/NA issues
tax_mat_char <- apply(tax_mat, 2, as.character)

# For each row (i.e., OTU), find the most specific (rightmost non-NA) taxon
finest_taxa <- apply(tax_mat_char, 1, function(x) {
  last_non_na <- tail(x[!is.na(x) & x != ""], 1)
  if (length(last_non_na) == 0) "Unclassified" else last_non_na
})

# Combine OTU ID and finest taxon
new_taxa_names <- paste(taxa_names(seqdata_RARE), "-", finest_taxa)

# Apply new names
taxa_names(seqdata_RARE) <- new_taxa_names
```

## Create temp factor levels
```{r}
seqdata_RARE@sam_data$burn_temp_C <- factor(seqdata_RARE@sam_data$burn_temp_C, 
                                   levels = c(20, 45, 75, 105, 150), ordered = TRUE)
```

# Data analysis

## Plotting

### Ordination (NMDS)

Using default of bray for distance (bray is default method for ordinate function)

####plotting nmds
```{r message=FALSE, warning=FALSE}
custom_colors <- c("#008080", "#52CCCC", "#FFE820", "orange", "#A80000")

ORDtotalvals<-ordinate(seqdata_RARE, "NMDS")

ORD <- plot_ordination(seqdata_RARE, ORDtotalvals, color = "burn_temp_C", shape = "burn_status") +
  scale_color_manual(values = custom_colors) +  # Custom colors
  stat_ellipse(aes(group = burn_status), level = 0.95, color = "orchid4", linewidth = .75, alpha=0.75)+
  geom_point(size = 3.5) +  # Increase point size
  labs(color = "Burn Temp (C)", shape = "Burn History") +  # Rename legend titles
  theme_minimal()+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.border = element_rect(fill = NA, colour = "grey70", linewidth = 1))

ORD
```

####saving ord plot
```{r}
#ggsave("figures/publication/STUDY_bray_ordination.jpeg", plot = ORD, width = 6, height = 4, dpi = 300)
```


### Rank abundance curves

#### Subsetting data
```{r}

```

#### Barplots
```{r}
plot_bar(seqdata_RARE, "Phylum", facet_grid = burn_status ~ burn_temp_C)
```


## Stats

###Converting to data frames
```{r}
sam_data_df <- as.data.frame(as.matrix(sample_data(seqdata_RARE)))
sam_data_df <- sam_data_df %>% 
  filter(!rownames(.) %in% c("N-N-D1", "N-N-D3"))

otu_table_df <- as.data.frame(t(otu_table(seqdata_RARE)))
otu_table_df <- otu_table_df %>% 
  filter(!rownames(.) %in% c("N-N-D1", "N-N-D3"))
```

### PERMANOVA

#### attach packages
```{r}
library(modelr)
library(broom)
library(AICcPermanova)
```

####burn status and burn temp
```{r}
status_temp_model <- adonis2(otu_table_df ~ burn_status*burn_temp_C, data=sam_data_df, perm=9999, by="terms")
status_temp_model
summary(status_temp_model)

AICc_permanova2(status_temp_model)
```

#### detach packages
```{r}
detach_package(modelr)
detach_package(broom)
detach_package(AICcPermanova)
```

### Pairwise adonis2 (PERMANOVA)

#### open packages
```{r}
library(devtools)

##installation code:
##Run: install_github("pmartinezarbizu/pairwiseAdonis/pairwiseAdonis")
library(pairwiseAdonis)
```

#### Creating interaction terms
```{r}
pair_sam_data_df <- sam_data_df %>% 
  mutate(burn_group = interaction(burn_status, burn_temp_C))
```

#### Creating distance matrix
```{r}
# Compute a distance matrix
bray_dist <- vegdist(otu_table_df, method = "bray")
```

#### Running pairwise adonis -- Burn status * Burn temp
```{r}
pairwise.adonis2(bray_dist ~ burn_group, pair_sam_data_df)
```




### TITAN

#### attach packages
```{r}
library(TITAN2)
```

####All samples

##### All samples -- Prepare data for TITAN
```{r}
##filtering rare species from OTU table to help TITAN run
seqdata_titan <- genefilter_sample(seqdata_RARE, filterfun_sample(function(x) x > 3), A=0.15*nsamples(seqdata_RARE))
      
seqdata_titan <- prune_taxa(seqdata_titan, seqdata_RARE)

##Extracting otu table
seqdf_titan <- as.data.frame(t(as(otu_table(seqdata_titan), "matrix")))

##Remove extraction controls
seqdf_titan <- seqdf_titan %>% 
  filter(!rownames(.) %in% c("N-N-D1", "N-N-D3"))

##Remove NAs
seqdf_titan <- na.omit(seqdf_titan)
sam_data_df <- na.omit(sam_data_df)
```

##### All samples -- burn temp
```{r}
##Making burn temp numeric
sam_data_df$burn_temp_C <- as.numeric(sam_data_df$burn_temp_C)

##Remove comment below to run titan (takes a few mins)
#all_temp_titan <- titan(env = sam_data_df$burn_temp_C, txa = seqdf_titan, pur.cut = 0.90)

all_temp_titan$sppmax %>%
as.data.frame() %>%
dplyr::select(zenv.cp, freq, maxgrp, IndVal, purity, reliability, filter) %>%
  utils::head()

## Plotting titan outputs (only have positively responding taxa)
all_temp_titan$sumz.cp
head(all_temp_titan$sppmax)
plot_sumz_density(all_temp_titan)
plot_taxa_ridges(all_temp_titan, axis.text.y = 10, xlabel = expression(paste("Burn Temp (C)")), rel_heights = c(6,2.5), xlim = c(10, 115))
dev.off() 
plot_taxa(all_temp_titan, xlabel = "Burn Temp (C)")
plot_cps(all_temp_titan, taxa.dist = FALSE, xlabel = "Burn Temp (C)", stacked=F)
```


####Burned only

##### Burned samples -- subset dfs
```{r}
# Subset sample metadata
sam_data_burned <- sam_data_df %>% filter(burn_status == "Burned")

# Subset OTU table to match the filtered sample IDs
seqdata_burned <- subset_samples(seqdata_RARE, sample_names(seqdata_RARE) %in% rownames(sam_data_burned))
```

##### Burned samples -- Prepare data for TITAN
```{r}
##filtering rare species from OTU table to help TITAN run
seqdata_burned_titan <- genefilter_sample(seqdata_burned, filterfun_sample(function(x) x > 3), A=0.15*nsamples(seqdata_burned))
      
seqdata_burned_titan <- prune_taxa(seqdata_burned_titan, seqdata_burned)

##Extracting otu table
seqdf_burned_titan <- as.data.frame(t(as(otu_table(seqdata_burned_titan), "matrix")))

##Remove extraction controls
seqdf_burned_titan <- seqdf_burned_titan %>% 
  filter(!rownames(.) %in% c("N-N-D1", "N-N-D3"))

##Remove NAs
seqdf_burned_titan <- na.omit(seqdf_burned_titan)
sam_data_burned <- na.omit(sam_data_burned)
```

##### Burned samples -- burn temp
```{r}
##Making burn temp numeric
sam_data_burned$burn_temp_C <- as.numeric(sam_data_burned$burn_temp_C)

##Remove comment below to run titan (takes a few mins)
#burned_temp_titan <- titan(env = sam_data_burned$burn_temp_C, txa = seqdf_burned_titan, pur.cut = 0.90)

burned_temp_titan$sppmax %>%
as.data.frame() %>%
dplyr::select(zenv.cp, freq, maxgrp, IndVal, purity, reliability, filter) %>%
  utils::head()

## Plotting titan outputs (only have positively responding taxa)
burned_temp_titan$sumz.cp
head(burned_temp_titan$sppmax)
plot_sumz_density(burned_temp_titan)
plot_taxa_ridges(burned_temp_titan, axis.text.y = 10, xlabel = expression(paste("Burn Temp (C)")), rel_heights = c(6,2.5), xlim = c(10, 115))
dev.off() 
plot_taxa(burned_temp_titan, xlabel = "Burn Temp (C)")
plot_cps(burned_temp_titan, taxa.dist = FALSE, xlabel = "Burn Temp (C)", stacked=F)
```




####Unburned only

##### Unburned samples -- subset dfs
```{r}
# Subset sample metadata
sam_data_unburned <- sam_data_df %>% filter(burn_status == "Unburned")

# Subset OTU table to match the filtered sample IDs
seqdata_unburned <- subset_samples(seqdata_RARE, sample_names(seqdata_RARE) %in% rownames(sam_data_unburned))
```

##### Unburned samples -- Prepare data for TITAN
```{r}
##filtering rare species from OTU table to help TITAN run
seqdata_unburned_titan <- genefilter_sample(seqdata_unburned, filterfun_sample(function(x) x > 3), A=0.15*nsamples(seqdata_unburned))
      
seqdata_unburned_titan <- prune_taxa(seqdata_unburned_titan, seqdata_unburned)

##Extracting otu table
seqdf_unburned_titan <- as.data.frame(t(as(otu_table(seqdata_unburned_titan), "matrix")))

##Remove extraction controls
seqdf_unburned_titan <- seqdf_unburned_titan %>% 
  filter(!rownames(.) %in% c("N-N-D1", "N-N-D3"))

##Remove NAs
seqdf_unburned_titan <- na.omit(seqdf_unburned_titan)
sam_data_unburned <- na.omit(sam_data_unburned)
```

##### Unburned samples -- burn temp
```{r}
##Making burn temp numeric
sam_data_unburned$burn_temp_C <- as.numeric(sam_data_unburned$burn_temp_C)

##Remove comment below to run titan (takes a few mins)
unburned_temp_titan <- titan(env = sam_data_unburned$burn_temp_C, txa = seqdf_unburned_titan, pur.cut = 0.90)

unburned_temp_titan$sppmax %>%
as.data.frame() %>%
dplyr::select(zenv.cp, freq, maxgrp, IndVal, purity, reliability, filter) %>%
  utils::head()

## Plotting titan outputs (only have positively responding taxa)
unburned_temp_titan$sumz.cp
head(unburned_temp_titan$sppmax)
plot_sumz_density(unburned_temp_titan)
plot_taxa_ridges(unburned_temp_titan, axis.text.y = 10, xlabel = expression(paste("Burn Temp (C)")), rel_heights = c(6,2.5), xlim = c(10, 115))
dev.off() 
plot_taxa(unburned_temp_titan, xlabel = "Burn Temp (C)")
plot_cps(unburned_temp_titan, taxa.dist = FALSE, xlabel = "Burn Temp (C)", stacked=F)
```




#### detach packages
```{r}
detach_package(TITAN2)
```

##Diversity

###Betadispersion
```{r}
z <- betadiver(otu_table_df, "z")
mod <- with(sam_data_df, betadisper(z, burn_status))
boxplot(mod)
modaov<-aov(mod$distances~mod$group)
summary(modaov)
```


### Beta diversity: Nestedness vs turnover
#### Burned samples only -- burn temp

##### Burned samples -- Subsetting samples
```{r}
burned_sam_data_df <- sam_data_df %>% 
  filter(burn_status == "Burned")

burned_otu_table_df <- otu_table_df %>%
  filter(rownames(.) %in% rownames(burned_sam_data_df))
```

##### Burned samples -- Creating distance matrix
```{r}
library(betapart)

##Turning df into a presence/absence matrix
burned_dfPA <- ifelse(burned_otu_table_df>0,1,0)

#Calculating community distances based on jaccard index(presence/abundance vs abundance)
burned_dist <- beta.pair(burned_dfPA, index.family = "jaccard")

#create a distance matrix for Sample Age
burned_daydist <- dist(burned_sam_data_df$burn_temp_C)
```

##### Burned samples -- Nestedness
```{r}
mantel(burned_dist$beta.jne, burned_daydist)
```

##### Burned samples -- Turnover 
```{r}
#statisical test to compare turnover across sampling day
mantel(burned_dist$beta.jtu, burned_daydist)
```

#### Unburned samples only -- burn temp

##### Unburned samples -- Subsetting samples
```{r}
unburned_sam_data_df <- sam_data_df %>% 
  filter(burn_status == "Unburned")

unburned_otu_table_df <- otu_table_df %>%
  filter(rownames(.) %in% rownames(unburned_sam_data_df))
```

##### Unburned samples -- Creating distance matrix
```{r}
library(betapart)

##Turning df into a presence/absence matrix
unburned_dfPA <- ifelse(unburned_otu_table_df>0,1,0)

#Calculating community distances based on jaccard index(presence/abundance vs abundance)
unburned_dist <- beta.pair(unburned_dfPA, index.family = "jaccard")

#create a distance matrix for Sample Age
unburned_daydist <- dist(unburned_sam_data_df$burn_temp_C)
```

##### Unburned samples -- Nestedness
```{r}
mantel(unburned_dist$beta.jne, unburned_daydist)
```

##### Unburned samples -- Turnover 
```{r}
#statisical test to compare turnover across sampling day
mantel(unburned_dist$beta.jtu, unburned_daydist)
```

#### Unheated (20 C) samples only -- burn temp

##### Unheated samples -- Subsetting samples
```{r}
unheated_sam_data_df <- sam_data_df %>% 
  filter(burn_temp_C == "20")

unheated_otu_table_df <- otu_table_df %>%
  filter(rownames(.) %in% rownames(unheated_sam_data_df))
```

##### Unheated samples -- Creating distance matrix
```{r}
library(betapart)

##Turning df into a presence/absence matrix
unheated_dfPA <- ifelse(unheated_otu_table_df>0,1,0)

#Calculating community distances based on jaccard index(presence/abundance vs abundance)
unheated_dist <- beta.pair(unheated_dfPA, index.family = "jaccard")

#convert to burn status to numeric
dist_unheated_sam_data_df <- unheated_sam_data_df %>% 
  mutate(burn_binary = ifelse(burn_status == "Burned", 1, 0))

#create a distance matrix for burn status
unheated_daydist <- dist(dist_unheated_sam_data_df$burn_binary) ####Not working properly
```

##### Unheated samples -- Nestedness
```{r}
mantel(unheated_dist$beta.jne, unheated_daydist) ####Not working properly
```

##### Unheated samples -- Turnover 
```{r}
#statisical test to compare turnover across sampling day
mantel(unheated_dist$beta.jtu, unheated_daydist) 
```


